---
title: "U.S. NOAA Storm Database Analysis"
author: "J. Baker"
date: "12/11/2020"
output:
  html_document: default
  pdf_document: default
---

**Introduction** 

In this project we explore the United States NOAA storm database.  The database
tracks major storm and weather event characteristics.  Characteristics include 
when and where storms occur as well as estimates of fatalities, injuries, property 
damage and loss.  The analysis below will focus on two questions and evaluate 
events that cause injury and fatalities and property damage.  

**Synopsis** 

Storm events reveal weather that is dangerous and that cause harm even death to 
those in its path.  It's revealed that populations and communities are effected 
significantly both in loss of human life and economically.  Floods and high winds 
resulted in BILLIONS of dollars in damages between the years 1950 and 2011.  The 
most significant impact (damages) were caused by drought.  Additional damages were 
caused by flood and hail events.  

**Loading**

*Loading Libraries* 

```{r warning=FALSE, message=FALSE}
# Loading required libraries 
library("ggplot2")
library("gridExtra")
library("R.utils")
```

*Loading and Reading Data*

```{r cache=TRUE}
#Reading dataset 
stormData <- read.csv("repdata_data_StormData.csv", sep = ",")
```

**Summary of Data**
*Review summary of the data that is available to process.*  
```{r}
summary(stormData)
```

**Data Processing** 
*We will process the data in the following ways for both Question 1 and 2 (See #):*
```{r}
#Trim data set to include required columns 
stormEvnt <- stormData[, c("BGN_DATE", "EVTYPE", "FATALITIES", "INJURIES", 
                            "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]

#Create a new subset for Q1 & Q2

#Selecting data for Fatalities and Injuries for Q1
eventHlth <- subset(stormEvnt, !stormEvnt$FATALITIES == 0 & !stormEvnt$INJURIES ==
                            0, select = c(EVTYPE, FATALITIES, INJURIES))

#Selecting data for Property Damage and Crop Damage for Q2
eventEcon <- subset(stormEvnt, !stormEvnt$PROPDMG == 0 & !stormEvnt$CROPDMG ==
                            0, select = c(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP))
```

**Question 1 - Across the US, which types of events (as indicated in the EVTYPE variable)
are most harmful with respect to population health?**                    

*Data Analysis for Question 1*

*We will process the data in the following ways (See #):* 
```{r}
#For Q1 - Create a separate data set for Injury and Fatalities 
eventHlth_Death <- aggregate(eventHlth$FATALITIES, by = list(eventHlth$EVTYPE),
                             FUN = sum)
#Naming columns 
colnames(eventHlth_Death) <- c("EVENTTYPE", "FATALITIES")
eventHlth_Inj <- aggregate(eventHlth$INJURIES, by = list(eventHlth$EVTYPE), 
                           FUN = sum)
colnames(eventHlth_Inj) <- c("EVENTTYPE", "INJURIES")
        
#Reorder 2 dataset and filter TOP 5 events for both 
eventHlth_Death <- eventHlth_Death[order(eventHlth_Death$FATALITIES, decreasing = TRUE),
][1:5, ]
eventHlth_Inj <- eventHlth_Inj[order(eventHlth_Inj$INJURIES, decreasing = TRUE),
                               ][1:5, ]
```

*Results for Question 1*

*We process data to include the TOP 5 major causes of both fatalities and injuries.* 
```{r}
#Fatalities plot stored in: Death_plot
Death_plot <- ggplot() + geom_bar(data = eventHlth_Death, aes(x = EVENTTYPE,
        y = FATALITIES, fill = interaction(FATALITIES, EVENTTYPE)), stat = "identity",
        show.legend = F) + theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
        xlab("Harmful Events") + ylab("No. of Fatalities") + ggtitle ("TOP 5 Weather Events Causing Fatalities") +
        theme(axis.text.x = element_text(angle = 30, hjust = 1))
#Injuries plot stored in: Inj_plot
Inj_plot <- ggplot() + geom_bar(data = eventHlth_Inj, aes(x = EVENTTYPE, y = INJURIES,
        fill = interaction(INJURIES, EVENTTYPE)), stat = "identity", show.legend = F) + 
        theme(axis.text.x = element_text(angle = 30, hjust = 1)) + xlab("Harmful Events") +
        ylab("No. of Injuries") + ggtitle("Top 5 Weather Events Causing Injuries") + 
        theme(axis.text.x = element_text(angle = 30, hjust = 1))

#Visualized and both plots generated.  Divide space equals two columns
grid.arrange(Death_plot, Inj_plot, ncol = 2)
```

*TORNADOS - The overwhelming cause of both FATALITIES and INJURIES related to weather events.*

**Question 2 - Across the United States, which types of events have the greatest economic consequences?** 

*Data Analysis for Question 2*
*We will process the data aimed at showing the most harmful events to an economy. (See #):* 
```{r}
#Selecting the entries for Economy
eventEcon <- subset(eventEcon, eventEcon$PROPDMGEXP == "K" | eventEcon$PROPDMGEXP == 
                            "k" | eventEcon$PROPDMGEXP == "M" | eventEcon$PROPDMGEXP == "m" |
                            eventEcon$PROPDMGEXP == "B" | eventEcon$PROPDMGEXP == "b")

eventEcon <- subset(eventEcon, eventEcon$CROPDMGEXP == "K" | eventEcon$CROPDMGEXP ==
                            "k" | eventEcon$CROPDMGEXP == "M" | eventEcon$CROPDMGEXP == "m" |
                            eventEcon$CROPDMGEXP == "B" | eventEcon$CROPDMGEXP == "b")

#Convert econ values to a number
eventEcon$PROPDMGEXP <- gsub("m", 1e+06, eventEcon$PROPDMGEXP, ignore.case = TRUE)

eventEcon$PROPDMGEXP <- gsub("k", 1000, eventEcon$PROPDMGEXP, ignore.case = TRUE)

eventEcon$PROPDMGEXP <- gsub("b", 1e+09, eventEcon$PROPDMGEXP, ignore.case = TRUE)

eventEcon$PROPDMGEXP <- as.numeric(eventEcon$PROPDMGEXP)

eventEcon$CROPDMGEXP <- gsub("m", 1e+06, eventEcon$CROPDMGEXP, ignore.case = TRUE)

eventEcon$CROPDMGEXP <- gsub("k", 1000, eventEcon$CROPDMGEXP, ignore.case = TRUE)

eventEcon$CROPDMGEXP <- gsub("b", 1e+09, eventEcon$CROPDMGEXP, ignore.case = TRUE)

eventEcon$CROPDMGEXP <- as.numeric(eventEcon$CROPDMGEXP)

eventEcon$PROPDMGEXP <- as.numeric(eventEcon$PROPDMGEXP)

#Sum of damage by each event type
eventEcon$TotalDMG <- (eventEcon$CROPDMG * eventEcon$CROPDMGEXP) + 
        (eventEcon$PROPDMG * eventEcon$CROPDMGEXP)

eventEcon <- aggregate(eventEcon$TotalDMG, by = list(eventEcon$EVTYPE),
                       FUN = sum)

colnames(eventEcon) <- c("EVTYPE", "TOTALDMG")

#Event type by rank (highest damage cost) using TOP 5 cols
eventEcon <- eventEcon[order(eventEcon$TOTALDMG, decreasing = TRUE),
                       ]
eventEcon <- eventEcon[1:5, ]
```

*Results for Question 2*
*We plot the graph to show the output.* 
```{r}
ggplot() + geom_bar(data = eventEcon, aes(x = EVTYPE, y = TOTALDMG, fill = interaction(TOTALDMG, EVTYPE)),
                    stat = "identity", show.legend = F) + theme(axis.text.x = element_text(angle = 30, hjust = 1)) +
                        xlab("Event Type") + ylab("Total Damage")
```

*The graph shows that FLOODS are the weather event that causes the most economic damage to a US economy.* 

**Summary**
*The data show that TORNADOs have the greatest effect on US population health and FLOODs have the greatest effect and severe consequences on the economy.*  